What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2022-12-30 paper/main.tex
png
update on examples

\begin{definition}[Linear Temporal Logic (LTL)]
    % cite Kristin's survey paper
    The syntax of an LTL formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi
    \end{align*}\label{ltl-defn}
\end{definition}
Intuitively, $\always \Phi$ says that formula $\Phi$ is true at every time step; $\eventually \Phi$ says that formula $\Phi$ is true either now or at sometime in the future time step; and $\nextt \Phi$ says formula $\Phi$ is true at the next time step immediately after the current one.
\begin{remark}
    The definition of \ltl\ can be extended to allow release and until as shown below, but they are definable using connectives in \cref{ltl-defn}, hence they are omitted in the definition. The tool we present in this paper, however, does support those operators.
    \begin{align*}
        \Phi ::= \cdots \mid \Phi \strongrelease \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi \mid \Phi \weakuntil \Phi
    \end{align*}
\end{remark}

\begin{definition}[Semantics of \ltl\ formula]
    The semantics of \ltl\ formula is defined via judgement $\pi, i \models \Phi$ where $\pi : \omega \rightarrow 2^{\AP}$ is the computation that stores the truthhood and falsehood of every atomic proposition at every time step, and $\omega$ is the set of natural numbers that denote the time step. The semantics follows exactly from \cite{Roz11}.
\end{definition}

@@

\begin{example}
    In \cref{fig:ex15}, one can reason about one timeline: the atomic proposition $a$ is false for finitely many time steps as signified by two nodes with $!a$ separated by $\cdots$ (could also be zero time step); followed by one node with $a$ which substantiates the specification of ``$a$ eventually true''; once $a$ is true at some point, the later time steps can behave arbitrarily as signified by the infinite run of $1$ (true) in the grey box.
    \begin{figure}[!h]
        \centering
        \includegraphics[scale=0.4]{examples/ex15/ex15.gv.png}
        \caption{Example timeline for specification ``$a$ is eventually true''}
        \label{fig:ex15}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\begin{example}
    Spot~\cite{SPOT-online} presents command-line tool for random \ltl\ formula generator \texttt{randltl}~\cite{duret.13.atva}. The following \ltl\ formula is generated from this tool.
    \[
    p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))
    \]
    This formula is reasonably complicated, and hard for human to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex13/ex13.png}
        \caption{Timeline for $p_2 \land (\eventually (\always p_0 \strongrelease  1) \weakrelease \nextt(\always p_1 \land ((p_0 \liff p_2) \stronguntil \eventually p_0)))$}
        \label{fig:ex13}
    \end{figure}
\end{example}
\begin{remark}
    These two examples show that our tool can generate reasonably clear diagrams for both real-world formula and randomly generated formula.
\end{remark}

@@

\section{Future work}
Based on our results, we raise the following questions as future works:
\begin{itemize}
    \item Can there be more simplifications methods as described in Section \ref{regex-simplify}? The answer is definitely positive. On one hand, more rule-based simplifications can be discovered by finding more patterns in the generated $\omega$-regular expressions. On the other hand, other simplification are possible: for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accepts the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduce the problem to equivalence test on $\omega$-regular expressions, which is decidable. In the similar vein, many more simplifications can be used, but the question is how efficient are they? In the end, one have to balance trade-off between the efficiency of the tool and the simplicity of the result.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulae with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timeline more formally so that one can do this process reversely (i.e. a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be a very useful tool in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

>>>>

\begin{definition}[Linear Temporal Logic (LTL)]
    % cite Kristin's survey paper
    The syntax of an LTL formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
    \end{align*}\label{ltl-defn}
\end{definition}

\begin{remark}
    Intuitively, $\always \Phi$ says that formula $\Phi$ is true at every time step; $\eventually \Phi$ says that formula $\Phi$ is true either now or at sometime in the future time step; and $\nextt \Phi$ says formula $\Phi$ is true at the next time step immediately after the current one; $\Phi \stronguntil \Psi$ says that formula $\Phi$ is true \textit{until} such time formula $\Psi$ becomes true; $\Phi \weakrelease \Psi$ says that formula $\Psi$ must be true now and remain true until formula $\Phi$ becomes true, after which point \textit{release} $\Psi$ (if formula $\Phi$ is never true, then formula $\Psi$ must remain true at all time).
\end{remark}

\begin{remark}
    The tool we provide in this paper uses a set of concrete syntax to represent those logical connectives, in order to avoid the difficulty of typesetting connectives such as $\always$ and $\eventually$. The concrete syntax is defined in \ref{sec:concrete-syntax}.
\end{remark}

\begin{definition}[Semantics of \ltl\ formula]
    The semantics of \ltl\ formula is defined via judgement $\pi, i \models \Phi$ where $\pi : \omega \rightarrow 2^{\AP}$ is the computation that stores the truthhood and falsehood of every atomic proposition at every time step, and $\omega$ is the set of natural numbers that denote the time step. The semantics follows exactly from Rozier~\cite{Roz11}.
\end{definition}

@@

\begin{example}
    In \cref{fig:ex15}, one can reason about one timeline: the atomic proposition $a$ is false for finitely many time steps as signified by two nodes with $!a$ separated by $\cdots$ (could also be zero time step); followed by one node with $a$ which substantiates the specification of ``$a$ eventually true''; once $a$ is true at some point, the later time steps can behave arbitrarily as signified by the infinite run of $1$ (true) in the grey box.
    \begin{figure}[!h]
        \centering
        \includegraphics[scale=0.4]{examples/ex15/ex15.png}
        \caption{Example timeline for specification ``$a$ is eventually true''}
        \label{fig:ex15}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\begin{example}
    Spot~\cite{SPOT-online} presents command-line tool for random \ltl\ formula generator \texttt{randltl}~\cite{duret.13.atva}. The following \ltl\ formula is generated from this tool.
    \[
    p_2 \land (\eventually \always p_0 \weakrelease \nextt(\always p_1 \land ((p_0 \limplies p_2 \land p_2 \limplies p_0) \stronguntil \eventually p_0)))
    \]
    This formula is reasonably complicated, and hard for human to reason about directly. For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex13}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex13/ex13.png}
        \caption{Timeline for $p_2 \land (\eventually \always p_0 \weakrelease \nextt(\always p_1 \land ((p_0 \limplies p_2 \land p_2 \limplies p_0) \stronguntil \eventually p_0)))$}
        \label{fig:ex13}
    \end{figure}
\end{example}
\begin{remark}
    These two examples show that our tool can generate reasonably clear diagrams for both real-world formula and randomly generated formula.
\end{remark}

@@

\section{Discussion}
Based on our results, we propose the following questions for discussions and future works:
\begin{itemize}
    \item Can there be more simplifications methods as described in Section \ref{regex-simplify}? The answer is definitely positive. On one hand, more rule-based simplifications can be discovered by finding more patterns in the generated $\omega$-regular expressions. On the other hand, other simplification are possible: for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accepts the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduce the problem to equivalence test on $\omega$-regular expressions, which is decidable. In the similar vein, many more simplifications can be used, but the question is how efficient are they? In the end, one have to balance trade-off between the efficiency of the tool and the simplicity of the result.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulae with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timeline more formally so that one can do this process reversely (i.e. a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be a very useful tool in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

\appendix
\section{Concrete Syntax of \texttt{ltl2timeline}} \label{sec:concrete-syntax}
Recall from \ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \texttt{ltl2timeline}, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic proposition} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]
